---
title: "The change of affordability and availability in the housing market for young adults in the Netherlands from 2005 until 2025."
author: "Björn Hildering 2887258, Tim Knegt 2882977, Willem van Beuningen 2787054, Robin Versnel, Seeger Tromp, Jaydane Levant 2851422"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
install.packages("readr")
install.packages("dplyr")
install.packages("ggplot2")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
require(readr)
require(dplyr)
require(ggplot2)
```


---
title: "Qualifying a social problem - The change of affordability and availability in the housing market for young adults in the Netherlands from 2005 until 2025"
subtitle: "Course: CPR01 - NL | Docent: Chantal Schouwenaar"
author:
  - "Björn Hildering (2887258)"
  - "Tim Knegt (2882977)"
  - "Willem van Beuningen (2787054)"
  - "Robin Versnel"
  - "Seeger Tromp (2889983)"
  - "Jaydane Levant (2851422)"
date: "23-06-2026"
output: pdf_document
---

# Part 1 - Identify a Social Problem
The Dutch housing market has changed significantly over the past two decades. For young adults, finding suitable housing has become increasingly difficult due to rising house prices, higher rents, and a continuing housing shortage (Nederlands Jeugdinstituut, n.d.). As a result, many young people struggle to enter the housing market.

This research examines how housing affordability and availability for young adults in the Netherlands changed between 2005 and 2025. It focuses on trends in house prices, rental costs, incomes and housing shortages to understand the main developments during this period.

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

Include the following:

-   Why is this relevant?

-   ...

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}

```

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)

```{r}

#Here we modify the dataset Inflatie
inflatie <- read.csv("data/Inflatie.csv")
inflatie <- inflatie %>%
  rename(Jaartal = Perioden) %>%
  rename(Inflatie = Jaarmutatie.CPI....) %>%
  select(Jaartal,Inflatie)
write.csv(inflatie, "data2.0/Inflatie_nieuw.csv", row.names = FALSE)
```

```{r}
#Here we modify the dataset Welvaart_van_huishoudens__kerncijfers
welvaart <- read.csv("data/Welvaart_van_huishoudens__kerncijfers.csv")
welvaart <- welvaart %>%
  rename(leeftijd_kostwinner = Kenmerken.van.huishoudens) %>%
  rename(Jaartal = Perioden) %>%
  rename(Gemiddeld_gestandaardiseerd_inkomen_x1000_euro = Gemiddeld.gestandaardiseerd.inkomen..1.000.euro.)
head(welvaart)
write.csv(welvaart, "data2.0/Welvaart_van_huishoudens_kerncijfers_nieuw.csv", row.names = FALSE)
```


```{r}
#Here we modify the dataset Huizenprijzen_Huizenvoorraad
huizenvoorraad <-read.csv("data/Huizenprijzen_Huizenvoorraad.csv")
huizenvoorraad <- huizenvoorraad %>%
  rename(Jaartal = Perioden) %>%
  rename(Provincies = Regio.s) %>%
  rename(Woningvoorraad_1_januari = Bouwen.en.wonen.Woningvoorraad.Voorraad.op.1.januari..aantal.)%>%
  rename(Nieuwbouw = Bouwen.en.wonen.Woningvoorraad.Nieuwbouw..aantal.) %>%
  select(Jaartal, Provincies, Woningvoorraad_1_januari, Nieuwbouw)
  
  write.csv(huizenvoorraad, "data2.0/Huizenprijzen_Huizenvoorraad_nieuw.csv", row.names = FALSE)
```

```{r}
#Here we modify the dataset bruto_en_besteedbaar_inkomen_leeftijdsgroep
inkomen <- read.csv("data/bruto_en_besteedbaar_inkomen_leeftijdsgroep.csv")
inkomen <- inkomen %>%
  slice(-1) %>%
  rename(Inkomensbronnen = V1) %>%
  rename(Jaren = V3) %>%
  rename(Kenmerken_van_huishoudens = V2) %>%
  rename(Particuliere_huishoudens_x1000 = V4) %>%
  rename(Totaal_inkomen_mln.euro = V5) %>%
  rename(Gemiddeld_inkomen_1000.euro = V6) %>%
  rename(Mediaan_inkomen_1000.euro = V7) %>%
  select(Inkomensbronnen, Jaren, Kenmerken_van_huishoudens, Particuliere_huishoudens_x1000, Totaal_inkomen_mln.euro, Gemiddeld_inkomen_1000.euro, Mediaan_inkomen_1000.euro)
write.csv(inkomen, "data2.0/bruto_en_besteedbaar_inkomen_leeftijdsgroep_nieuw.csv", row.names = FALSE)
```

```{r}
#Here we modify the dataset Bestaand_koopwoningen verkoopprijzen
Bestaande_koopwoningen_verkoopprijzen_woningtype <- read.csv("data/Bestaande_koopwoningen_verkoopprijzen_woningtype.csv")
Bestaande_koopwoningen_verkoopprijzen_woningtype <- Bestaande_koopwoningen_verkoopprijzen_woningtype %>%
  rename( Jaartal = Perioden)
Bestaande_koopwoningen_verkoopprijzen_woningtype <- Bestaande_koopwoningen_verkoopprijzen_woningtype %>%
  rename(Aantal_verkochte_woningen = Verkochte.bestaande.koopwoningen.Verkochte.woningen..aantal.)
Bestaande_koopwoningen_verkoopprijzen_woningtype <- Bestaande_koopwoningen_verkoopprijzen_woningtype %>%
  rename( Gemiddelde_verkoopprijs_euro = Gemiddelde.verkoopprijs..euro.)
Bestaande_koopwoningen_verkoopprijzen_woningtype <- Bestaande_koopwoningen_verkoopprijzen_woningtype %>%
  select(Jaartal, Aantal_verkochte_woningen, Gemiddelde_verkoopprijs_euro)
  write.csv(Bestaande_koopwoningen_verkoopprijzen_woningtype, "data2.0/Bestaande_koopwoningen_verkoopprijzen_woningtype_nieuw.csv", row.names = FALSE)
```


```{r}
#Here we modify the dataset bevolkingsgroei to only keep the relevant information and use the right years.
bevolkingsgroei <-read.csv("data/Bevolkingsgroei_opbouw_2005_2019.csv")
bevolkingsgroei <- bevolkingsgroei %>%
  rename( jong_volwassenen_20_tot_45_jaar = Bevolking.op.1.januari.Naar.leeftijd.20.tot.45.jaar..x.1.000.)

bevolkingsgroei <- bevolkingsgroei%>%
  transmute(jong_volwassenen_20_tot_45_jaar)

bevolkingsgroei <- bevolkingsgroei %>%
  mutate(Jaar = 2005:2019)
bevolkingsgroei <- bevolkingsgroei %>%
  relocate(Jaar)
write.csv(bevolkingsgroei, "data2.0/Bevolkingsgroei_nieuw.csv", row.names = FALSE)
```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

``` r
inline_code = TRUE
```

```{r}
#Here we modify the dataset Huizenprijzen_verschillende_woningtypes
huizenprijzen <- read_csv("data/Huizenprijzen_verschillende_woningtypes.csv")
names(huizenprijzen)[names(huizenprijzen) == "Perioden"] <- "Jaren"
names(huizenprijzen)[names(huizenprijzen) == "Type.woning"] <- "Woningtype"
names(huizenprijzen)[names(huizenprijzen) == "Prijsindex.bestaande.koopwoningen.Prijsindex.verkoopprijzen..2020.100."] <- "Verkoopprijzen.bestaande.koopwoningen"
names(huizenprijzen)[names(huizenprijzen) == "Prijsindex.bestaande.koopwoningen.Ontwikkeling..t.o.v..een.jaar.eerder...."] <- "Verkoopprijzen.tov.voorgaand.jaar"
write.csv(huizenprijzen, "data2.0/nieuw_Huizenprijzen_per_woningtype.csv" , row.names = FALSE)
```


These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning}
#Start joining dataset inflatie_nieuw and Welvaart_van_huishoudens_kerncijfers_nieuw
welvaart <- read_csv("data/Welvaart_van_huishoudens_kerncijfers_nieuw.csv")
welvaart <- welvaart %>%
  mutate(Jaartal = as.numeric(Jaartal))
inflatie <- read_csv("data/Inflatie_nieuw.csv")
result <- left_join(welvaart, inflatie, by = "Jaartal")
head(result)

```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}

```

Variable 2

```{r gen_var2}

```

## 3.3 Visualize temporal variation

```{r}

```

## 3.4 Visualize spatial variation

```{r visualise_map}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation


Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
