---
title: "The change of affordability and availability in the housing market for young adults in the Netherlands from 2005 until 2025."
author: "Björn Hildering 2887258, Tim Knegt 2882977, Willem van Beuningen 2787054, Robin Versnel 2862644, Seeger Tromp, Jaydane Levant 2851422"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
install.packages("readr")
install.packages("dplyr")
install.packages("ggplot2")
install.packages(c(
  "geodata", "terra", 
  "janitor", "scales", "patchwork"))
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
require(readr)
require(dplyr)
require(ggplot2)
library(geodata)
library(terra)
library(janitor)
library(scales)
library(patchwork)

```

# Title Page

Include your names

Include the tutorial group number

Include your tutorial lecturer's name

# Part 1 - Identify a Social Problem
The Dutch housing market has changed significantly over the past two decades. For young adults, finding suitable housing has become increasingly difficult due to rising house prices, higher rents, and a continuing housing shortage (Nederlands Jeugdinstituut, n.d.). As a result, many young people struggle to enter the housing market.

This research examines how housing affordability and availability for young adults in the Netherlands changed between 2005 and 2025. It focuses on trends in house prices, rental costs, incomes and housing shortages to understand the main developments during this period.


Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem
The Dutch housing market has become increasingly inaccessible for young adults, making the move toward independent living far more difficult. This is socially relevant because stable and affordable housing is a foundation for financial independence, mental well‑being, and equal life opportunities. When young people cannot secure a home, they delay major life steps such as studying, working full‑time, or forming households—effects that ripple through society. The OECD notes that the Netherlands is among the countries where young adults face the greatest barriers to affordable housing due to rising prices and persistent shortages (OECD, 2023).

A combination of sharply increasing house prices, higher rents, and limited supply has created a structural imbalance between demand and availability. These pressures fall disproportionately on first‑time buyers and young renters, widening generational inequality: older households benefit from accumulated housing wealth, while younger cohorts struggle to enter the market at all.

Include the following:

-   Why is this relevant?

-   ...

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}

```


## 2.2 Provide a short summary of the dataset(s)

# trendgrafieken per provincie Huizenprijzen/voorraad

```{r}
library(tidyverse)
library(janitor)
library(scales)

df <- read_csv("data/Huizenprijzen_Huizenvoorraad_nieuw.csv") |> 
  clean_names()

df_index <- df |> 
  pivot_longer(
    cols = c(woningvoorraad_1_januari, nieuwbouw),
    names_to = "variabele",
    values_to = "waarde"
  ) |> 
  group_by(provincies, variabele) |> 
  arrange(jaartal, .by_group = TRUE) |> 
  mutate(index = waarde / first(waarde) * 100) |> 
  ungroup()

ggplot(df_index, aes(x = jaartal, y = index, color = variabele)) +
  geom_line(linewidth = 1) +
  facet_wrap(~ provincies, ncol = 3, scales = "free_y") +
  scale_x_continuous(
    breaks = seq(2005, 2025, by = 5),
    limits = c(2005, 2025)
  ) +
  labs(
    title = "Ontwikkeling woningvoorraad en nieuwbouw per provincie",
    subtitle = "Index: eerste beschikbare jaar = 100; waarden boven 100 betekenen groei, onder 100 afname",
    x = "Jaartal",
    y = "Index, eerste beschikbare jaar = 100",
    color = "Variabele"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```




```{r}

#Here we modify the dataset Inflatie
inflatie <- read.csv("data/Inflatie.csv")
inflatie <- inflatie %>%
  rename(Jaartal = Perioden) %>%
  rename(Inflatie = Jaarmutatie.CPI....) %>%
  select(Jaartal,Inflatie)
write.csv(inflatie, "data 2.0/Inflatie_nieuw.csv", row.names = FALSE)
```

```{r}
#Here we modify the dataset Welvaart_van_huishoudens__kerncijfers
welvaart <- read.csv("data/Welvaart_van_huishoudens__kerncijfers.csv")
welvaart <- welvaart %>%
  rename(leeftijd_kostwinner = Kenmerken.van.huishoudens) %>%
  rename(Jaartal = Perioden) %>%
  rename(Gemiddeld_gestandaardiseerd_inkomen_x1000_euro = Gemiddeld.gestandaardiseerd.inkomen..1.000.euro.)
head(welvaart)
write.csv(welvaart, "data 2.0/Welvaart_van_huishoudens_kerncijfers_nieuw.csv", row.names = FALSE)
```


```{r}
#Here we modify the dataset Huizenprijzen_Huizenvoorraad
huizenvoorraad <-read.csv("data/Huizenprijzen_Huizenvoorraad.csv")
huizenvoorraad <- huizenvoorraad %>%
  rename(Jaartal = Perioden) %>%
  rename(Provincies = Regio.s) %>%
  rename(Woningvoorraad_1_januari = Bouwen.en.wonen.Woningvoorraad.Voorraad.op.1.januari..aantal.)%>%
  rename(Nieuwbouw = Bouwen.en.wonen.Woningvoorraad.Nieuwbouw..aantal.) %>%
  select(Jaartal, Provincies, Woningvoorraad_1_januari, Nieuwbouw)
  
  write.csv(huizenvoorraad, "data 2.0/Huizenprijzen_Huizenvoorraad_nieuw.csv", row.names = FALSE)
```

```{r}
#Here we modify the dataset bruto_en_besteedbaar_inkomen_leeftijdsgroep
inkomen <- read.csv("data/bruto_en_besteedbaar_inkomen_leeftijdsgroep.csv")
inkomen <- inkomen %>%
  slice(-1) %>%
  rename(Inkomensbronnen = V1) %>%
  rename(Jaren = V3) %>%
  rename(Kenmerken_van_huishoudens = V2) %>%
  rename(Particuliere_huishoudens_x1000 = V4) %>%
  rename(Totaal_inkomen_mln.euro = V5) %>%
  rename(Gemiddeld_inkomen_1000.euro = V6) %>%
  rename(Mediaan_inkomen_1000.euro = V7) %>%
  select(Inkomensbronnen, Jaren, Kenmerken_van_huishoudens, Particuliere_huishoudens_x1000, Totaal_inkomen_mln.euro, Gemiddeld_inkomen_1000.euro, Mediaan_inkomen_1000.euro)
write.csv(inkomen, "data 2.0/bruto_en_besteedbaar_inkomen_leeftijdsgroep_nieuw.csv", row.names = FALSE)
```

```{r}
#Here we modify the dataset Bestaand_koopwoningen verkoopprijzen
Bestaande_koopwoningen_verkoopprijzen_woningtype <- read.csv("data/Bestaande_koopwoningen_verkoopprijzen_woningtype.csv")
Bestaande_koopwoningen_verkoopprijzen_woningtype <- Bestaande_koopwoningen_verkoopprijzen_woningtype %>%
  rename( Jaartal = Perioden)
Bestaande_koopwoningen_verkoopprijzen_woningtype <- Bestaande_koopwoningen_verkoopprijzen_woningtype %>%
  rename(Aantal_verkochte_woningen = Verkochte.bestaande.koopwoningen.Verkochte.woningen..aantal.)
Bestaande_koopwoningen_verkoopprijzen_woningtype <- Bestaande_koopwoningen_verkoopprijzen_woningtype %>%
  rename( Gemiddelde_verkoopprijs_euro = Gemiddelde.verkoopprijs..euro.)
Bestaande_koopwoningen_verkoopprijzen_woningtype <- Bestaande_koopwoningen_verkoopprijzen_woningtype %>%
  select(Jaartal, Aantal_verkochte_woningen, Gemiddelde_verkoopprijs_euro)
  write.csv(Bestaande_koopwoningen_verkoopprijzen_woningtype, "data 2.0/Bestaande_koopwoningen_verkoopprijzen_woningtype_nieuw.csv", row.names = FALSE)
```


```{r}
#Here we modify the dataset bevolkingsgroei to only keep the relevant information and use the right years.
bevolkingsgroei <-read.csv("data/Bevolkingsgroei_opbouw_2005_2019.csv")
bevolkingsgroei <- bevolkingsgroei %>%
  rename( jong_volwassenen_20_tot_45_jaar = Bevolking.op.1.januari.Naar.leeftijd.20.tot.45.jaar..x.1.000.)

bevolkingsgroei <- bevolkingsgroei%>%
  transmute(jong_volwassenen_20_tot_45_jaar)

bevolkingsgroei <- bevolkingsgroei %>%
  mutate(Jaar = 2005:2019)
bevolkingsgroei <- bevolkingsgroei %>%
  relocate(Jaar)
write.csv(bevolkingsgroei, "data 2.0/Bevolkingsgroei_nieuw.csv", row.names = FALSE)
```

```{r}
#Here we order the dataset to remove that it is grouped per function.
nieuwbouw <- read.csv("data 2.0/Huizenprijzen_Huizenvoorraad_nieuw.csv")

tabel_per_jaar <- nieuwbouw %>%
  group_by(Jaartal) %>%
  summarise(
    Woningvoorraad_1_januari = sum(Woningvoorraad_1_januari, na.rm = TRUE),
    Nieuwbouw = sum(Nieuwbouw, na.rm = TRUE)
  )
write.csv(tabel_per_jaar,"data 2.0/Huizenvoorraad_nieuw.csv")
```



``` r
inline_code = TRUE
```

```{r}
#Here we modify the dataset Huizenprijzen_verschillende_woningtypes
huizenprijzen <- read_csv("data/Huizenprijzen_verschillende_woningtypes.csv")
names(huizenprijzen)[names(huizenprijzen) == "Perioden"] <- "Jaren"
names(huizenprijzen)[names(huizenprijzen) == "Type.woning"] <- "Woningtype"
names(huizenprijzen)[names(huizenprijzen) == "Prijsindex.bestaande.koopwoningen.Prijsindex.verkoopprijzen..2020.100."] <- "Verkoopprijzen.bestaande.koopwoningen"
names(huizenprijzen)[names(huizenprijzen) == "Prijsindex.bestaande.koopwoningen.Ontwikkeling..t.o.v..een.jaar.eerder...."] <- "Verkoopprijzen.tov.voorgaand.jaar"
write.csv(huizenprijzen, "data 2.0/nieuw_Huizenprijzen_per_woningtype.csv" , row.names = FALSE)
```




## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning}
 
#Start joining dataset inflatie_nieuw and Welvaart_van_huishoudens_kerncijfers_nieuw
welvaart <- read_csv("data 2.0/Welvaart_van_huishoudens_kerncijfers_nieuw.csv")
welvaart <- welvaart %>%
  mutate(Jaartal = as.numeric(Jaartal))
inflatie <- read_csv("data 2.0/Inflatie_nieuw.csv")
result <- left_join(welvaart, inflatie, by = "Jaartal")
write.csv(result,"data 2.0/Samenvoeging1.csv" , row.names = FALSE)
```

```{r}
#Start joining dataset Samenvoeging1 and bruto_en_besteedbaar_inkomen_leeftijdsgroep_nieuw.csv
samenvoeging <- read_csv("data 2.0/Samenvoeging1.csv")
brutoinkomen <- read_csv("data 2.0/bruto_en_besteedbaar_inkomen_leeftijdsgroep_nieuw.csv")
brutoinkomen <- brutoinkomen %>%
  rename("Jaartal" = "Jaren")%>%
  mutate(Jaartal = as.numeric(Jaartal))
samenvoeging2 <- left_join(brutoinkomen, samenvoeging, by = "Jaartal")
write.csv(samenvoeging2, "data 2.0/Samenvoeging2.csv" , row.names = FALSE)

samenvoeging2$Jaartal[c(14,28)]<- 2024
samenvoeging2$Inflatie[c(14,28)]<- 3.3

```



Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}

```

Variable 2

```{r gen_var2}

```

## 3.3 Visualize temporal variation

```{r}

```

## 3.4 Visualize spatial variation

```{r visualise_map}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation


Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
